Fluent statistical computing interfaces for biological data analysis

### 2020-10-28 response/response.Rmd

---
title: Thesis Amendments
author: Stuart Lee
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
bibliography: response.bib
fontfamily: mathpazo
citation_package: biblatex
---

@@

> p. 82: “Like, when, using” -> Like when using “While our software, liminal is able “ -> While our software, liminal, is able

Amended.

>Section 5.3.1 “Finding Gestalt: focus and context” - this is probably a clever metaphor, but I didn’t understand it.

This was a reference to @Buja1996-fk paper mentioned in the paragraph above. Here "Gestalt" refers to Gestalt theory in perception and psychology, it refers to identifying patterns in visual forms.

@@

> p. 98: “We have shown in the case studies, that” -> no comma

Amended.

## Assistant Professor Stephanie C. Hicks {-}

Thank you for your thorough review and thoughtful suggestions. This feedback is extremely helpful and I am looking forward to incorporating your suggestions into future work.

### Addressing General Comments {-}

**On Scalability**

@@

I think the facet labels are sufficient in conjunction with the figure captions.

### Chapter 5 comments {-}

>  One question I had as I was reading throughout this chapter, was what is your recommendation for users who want to calculate distances? can the liminal framework be used to better approximate distances instead of users who perform clustering directly on top of the tSNE or UMAP spaces? While
is not ideal, and it is recommended clustering should performed in the PC space, it happens all the time. Another way of framing this question is can your framework be used to improve distance estimates in the tour with a NLDR method as you note it preserves global structure?

This is a really interesting question. To begin, I have amended the clustering case study to comment on a more realistic approach to perform cluster verification. I like the idea of using the tour to calibrate distances obtained from an NLDR method (or the other way around). As mentioned in the text if there are (well-behaved) clusters in the data, it appears that t-SNE will find them but often messes up their orientation and topology. I think you would get some of the way there using the spatial linking approach I described. Another approach would be to look at distance preservation between successive tour frames. One could then ask, how often are these groups of points close according to the tour? Of interest would also be the manual clustering approaches described in @Cook2007-ld. This is left as future work.

@@

> Pg 87: “inn” should be “in”

Amended.

> Pg 88: Inside Figure 5.1A and B, could you have liminal label which plot is which i.e. t-SNE vs tour
view. Also, could axis labels be added? Maybe also add this information to the Figure legend? – Pg 89 – same comment

I think this is not immediately obvious in the static figure but is obvious in the video and when using the **liminal** package that the embedding view is always on the left, and the tour view is on the right (it will be the view that is animated). I think the suggestion of titling the views is a good one and will place that into the software in future. Regarding axis labels, I don't think they make sense for the tour since it's really a linear combination of multiple variables (there's an option to switch on a biplot view for the tour). I am also reluctant to include them for t-SNE because they can be misleading for the reasons mentioned in the chapter, however I do think this would be good to have as an option for **liminal** in future.

### 2020-10-28 response/response.Rmd

Thank you for your thorough review and thoughtful suggestions. This feedback is extremely helpful and I am looking forward to incorporating your suggestions into future work.

### Addressing General Comments {-}

**On Scalability**

I would like to comment that I have avoided performance benchmarking for two main reasons. First, I believe it is out of scope for the thesis, since my focus has been on the development of programming and user interfaces. Second, my implementation of **plyranges** is deeply entwined with the underlying core Bioconductor infrastructure, especially the data structures and methods found in the **IRanges** and **GenomicRanges** packages. This would result in any benchmarking of **plyranges** turning into benchmarking of those packages which has been done elsewhere.

@@

Regarding out of memory data structures like `DelayedArray`, it is possible to incorporate them as columns to a `GRanges` object and then the full **plyranges** API would be available. Currently, there appear to be no plans in Bioconductor for out of memory `Ranges`, and indeed there are times where it is more appropriate to switch to a rectangular data layout for computation, as is reflected in the workflow described in chapter 3 of the thesis. I believe the development of a scalable out of memory interface for genomic data analysis will require the transformation between data measured along the genome (covered by the **plyranges** API) and the rectangular layouts that enable out of disk computing like `SummarizedExperiment`. This is an exciting research area that I would like to explore in the currently under development **plyexperiment**.

**On alignment vs. pseudoalignment**

I have amended the introduction to include a note about alignment. 
I believe it is out of scope to go into the finer details of when to use alignment or pseudo-alignment methods to obtain counts for RNA-seq data. I have included results from both approaches to highlight that **plyranges** is agnostic to the method used and is flexible for many types of data.

@@

This chapter has already been published and it appears most of the comments pertain to the use and learning of the **plyranges** package. I believe they are best addressed through the **plyranges** documentation and vignette. I have amended the thesis to include the package vignette as an appendix.

### Chapter 3 comments {-}

> Almost all of Chapter 2 discusses the use of GRanges objects. As most of Chapter 3 discusses the use of the SummarizedExperiment object, could you add a motivating figure at the start explaining the relationship between a SummarizedExperiment and GRanges object (i.e. rowRanges)? I do not think this is obvious to a reader who may not very familiar with these objects. However, I think it is an important aspect to explain in this chapter.

I agree and have amended this by adding a figure which explains what a `SummarizedExperiment`  is and how it is related to a `GRanges`.

@@

> Pg 40: I think it would be useful if you could comment on the results() function a bit more to explain
what is happening under the hood there. It’s a pretty critical step for everything downstream after identifying the DE genes in preparation for the integration step. As it stands, I felt like it was kind of glossed over.

Amended.

>Pg 45: More motivation could be provided on why you want to generate random samples of size equal to the number of DE-genes from the other_genes with your larger question in mind of “how many DA peaks are near DE genes relative to”other" non-DE genes¿‘. It is stated on pg 46”we minimize the variance on the enrichment statistics induced by the sampling process", but it might be good to explain this in greater detail at a higher level.

I have added a few more sentences to this section which describes the motivation for selecting sets of the same size between non-DE and DE genes, and why we want to do sub-sampling.

### 2020-10-29 Rmd/04-intron.Rmd
typo fixes
fix typos
clarify labelling

(ref:example-trace): An illustrative coverage trace plot from Chapter \@ref(ch:plyranges). The coverage _score_ is defined as the number of reads that overlap a single base position within a reference genome. For example, at position 21.807 of chromosome 8 there are approximately 12 reads overlapping it. One could also think of a coverage trace as a histogram with the bin width varies as you move along the genome.

@@

External reference annotations, perhaps transcripts or exons, 
can be coerced to _GRanges_ objects, are  incorporated into the coverage _GRanges_ by taking the intersection of
the annotation with the _GRanges_ using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
per base coverage that are overlapped the genomic features in the annotation,
along side any metadata about the features themselves. Since our main workflow
interest is in discovering coverage traces with IR profiles, **superintronic**
provides some syntactic sugar for unravelling gene annotations into their exonic
and intronic parts, and intersecting them with a coverage _GRanges_.

@@

The gene annotation files were
obtained as GFF files from RefSeq and used to construct the exonic
and intronic parts of each gene as _GRanges_ object using **superintronic**. 
We further filtered genes that had a single exon or genes that were overlapping
others in the annotation and that were not on the main contigs of the reference genome (i.e. excluding mitochondria) to simplify our analysis and reduce any coverage ambiguity. This left 18,270 genes available for 
computing data descriptors on in order to detecting IR-like coverage traces. Across each combination of genotype and line, we first $\log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene. For each gene,
the mean and standard deviation of the log-transformed coverage score weighted by the number of bases covered were computed over all intron and exon parts within the groupings of 
genotype and line. We also computed the bases above descriptor for intron
features. It refers to the total number of bases within an intron that has a score above the overall average exon coverage score. The scatter plot 
matrix view of these descriptors is shown in Figure \@ref(fig:si-splom)
for a single biological group in the experiment. Using these views and
by summarising over the data descriptors, we came up with thresholds
for finding genes that have IR-like traces within each biological group.

>>>>

(ref:example-trace) An illustrative coverage trace plot from Chapter \@ref(ch:plyranges). The coverage _score_ is defined as the number of reads that overlap a single base position within a reference genome. For example, at position 21.807 of chromosome 8 there are approximately 12 reads overlapping it.

@@

External reference annotations, perhaps transcripts or exons, 
can be coerced to _GRanges_ objects, are  incorporated into the coverage _GRanges_ by taking the intersection of
the annotation with the _GRanges_ using an overlap intersect join from the **plyranges** software. The resulting intersection will now contain the
per base coverage that are overlapped the genomic features in the annotation,
along side any metadata about the features themselves. Since our main workflow
interest is in discovering coverage traces with IR profiles, **superintronic**
provides some syntactic sugar for unravelling gene annotations into their exonic
and intronic parts, and intersecting them with a coverage _GRanges_. Split reads that cross the boundaries of exon and intron parts are counted towards both unless filtered beforehand.

@@

The gene annotation files were
obtained as GFF files from RefSeq as is standard for Zebrafish analyses. The GFF was used to construct the exonic
and intronic parts of each gene as _GRanges_ object using **superintronic**. 
We further filtered genes that had a single exon or genes that were overlapping
others in the annotation and that were not on the main contigs of the reference genome (i.e. excluding mitochondria) to simplify our analysis and reduce any coverage ambiguity. This left 18,270 genes available for 
computing data descriptors on in order to detecting IR-like coverage traces. Across each combination of genotype and line, we first $\log_2$ transformed the coverage score with an offset of one, and then intersected the coverage _GRanges_ with exonic and intronic features of each gene. For each gene,
the mean and standard deviation of the log-transformed coverage score weighted by the number of bases covered were computed over all intron and exon parts within the groupings of 
genotype and line. We also computed the bases above descriptor for intron
features. It refers to the total number of bases within an intron that has a score above the overall average exon coverage score. The scatter plot 
matrix view of these descriptors is shown in Figure \@ref(fig:si-splom)
for a single biological group in the experiment. Using these views and
by summarising over the data descriptors, we came up with thresholds
for finding genes that have IR-like traces within each biological group.

### 2020-10-29 Rmd/05-embeddings.Rmd
clarify labelling

Tours provide a supportive visualisation to NLDR graphics, and 
can be easily incorporated into an analysts workflow with our software package, 
**liminal**.  Our interface allows analysts to quickly compare 
views from embedding methods and see how an embedding method 
preserves or alters the geometry of their data. Using multiple concatenated and
linked views with the tour enhances interaction techniques, and allows 
analysts to perform  cluster orientation tasks via linked highlighting and 
brushing [@McDonald1982-ew; @Becker1987-gd]. 
This approach allows our interface to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "85%"}
display_screenshot("mouse-01", vid_lookup)
```

>>>>

Tours provide a supportive visualisation to NLDR graphics, and 
can be easily incorporated into an analysts workflow with our software package, 
**liminal**.  Our interface allows analysts to quickly compare 
views from embedding methods and see how an embedding method 
preserves or alters the geometry of their data. Using multiple concatenated and
linked views with the tour enhances interaction techniques, and allows 
analysts to perform  cluster orientation tasks via linked highlighting and 
brushing [@McDonald1982-ew; @Becker1987-gd]. 
This approach allows our interface to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.

@@

```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "90%"}
display_screenshot("mouse-01", vid_lookup)
```
